---
title: "Mulder&Hamaker(2021)Three Extensions of RI-CLPM"
author: "Yeon Suh"
date: "`r Sys.Date()`"
output: html_document
---
```{r}
library("lavaan")
```


#1 Basic Model

```{r}
RICLPM <- '
  # Create between components (random intercepts)
  RIx =~ 1*x1 + 1*x2 + 1*x3 + 1*x4 + 1*x5
  RIy =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5
  
  # Create within-person centered variables
  wx1 =~ 1*x1
  wx2 =~ 1*x2
  wx3 =~ 1*x3 
  wx4 =~ 1*x4
  wx5 =~ 1*x5
  wy1 =~ 1*y1
  wy2 =~ 1*y2
  wy3 =~ 1*y3
  wy4 =~ 1*y4
  wy5 =~ 1*y5

  # Estimate lagged effects between within-person centered variables
  wx2 + wy2 ~ wx1 + wy1
  wx3 + wy3 ~ wx2 + wy2
  wx4 + wy4 ~ wx3 + wy3
  wx5 + wy5 ~ wx4 + wy4

  # Estimate covariance between within-person centered variables at first wave
  wx1 ~~ wy1 # Covariance
  
  # Estimate covariances between residuals of within-person centered variables 
  # (i.e., innovations)
  wx2 ~~ wy2
  wx3 ~~ wy3
  wx4 ~~ wy4
  wx5 ~~ wy5
  
  # Estimate variance and covariance of random intercepts
  RIx ~~ RIx
  RIy ~~ RIy
  RIx ~~ RIy

  # Estimate (residual) variance of within-person centered variables
  wx1 ~~ wx1 # Variances
  wy1 ~~ wy1 
  wx2 ~~ wx2 # Residual variances
  wy2 ~~ wy2 
  wx3 ~~ wx3 
  wy3 ~~ wy3 
  wx4 ~~ wx4 
  wy4 ~~ wy4 
  wx5 ~~ wx5
  wy5 ~~ wy5
'
```


#2 Constraints Over Time
Imposing constraints to the model can be achieved through pre-multiplication. It means that we have to prepend the number that we want to fix the parameter to, and an asterisk, to the parameter in the model specification. For example, F =~ 0*x1 fixes the factor loading of item x1 to factor F to 0. Using pre-multiplication we can also constrain parameters to be the same by giving them the same label. Below we specify an RI-CLPM with the following constraints:

fixed auto-regressive and cross-lagged relations over time, wx2 ~ a*wx1 + b*wy1; ...,
time-invariant (residual) (co-)variances in the within-person part wx2 ~~ cov*wy2; ..., wx2 ~~ vx*wx2; ..., and wy2 ~~ vy*wy2; ..., and
constrained grand means over time, x1 + ... ~ mx*1 and y1 + ... ~ my*1.

```{r}
RICLPM5 <- '
  # Create between components (random intercepts)
  RIx =~ 1*x1 + 1*x2 + 1*x3 + 1*x4 + 1*x5
  RIy =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5
  
  # Create within-person centered variables
  wx1 =~ 1*x1
  wx2 =~ 1*x2
  wx3 =~ 1*x3 
  wx4 =~ 1*x4
  wx5 =~ 1*x5
  wy1 =~ 1*y1
  wy2 =~ 1*y2
  wy3 =~ 1*y3
  wy4 =~ 1*y4
  wy5 =~ 1*y5
  
  # Estimate lagged effects between within-person centered variables 
  # (constrained)
  wx2 ~ a*wx1 + b*wy1 
  wy2 ~ c*wx1 + d*wy1
  wx3 ~ a*wx2 + b*wy2
  wy3 ~ c*wx2 + d*wy2
  wx4 ~ a*wx3 + b*wy3
  wy4 ~ c*wx3 + d*wy3
  wx5 ~ a*wx4 + b*wy4
  wy5 ~ c*wx4 + d*wy4
  
  # Estimate covariances between residuals of within-person centered variables 
  # (i.e., innovations, constrained)
  wx2 ~~ cov*wy2
  wx3 ~~ cov*wy3
  wx4 ~~ cov*wy4 
  wx5 ~~ cov*wy5
  
  # Estimate covariance between within-person centered variables at first wave
  wx1 ~~ wy1 # Covariance
  
  # Estimate variance and covariance of random intercepts
  RIx ~~ RIx
  RIy ~~ RIy
  RIx ~~ RIy
  
  # Estimate (residual) variance of within-person centered variables 
  # (constrained)
  wx1 ~~ wx1 # Variance
  wy1 ~~ wy1 
  wx2 ~~ vx*wx2 # Residual variance
  wy2 ~~ vy*wy2 
  wx3 ~~ vx*wx3 
  wy3 ~~ vy*wy3 
  wx4 ~~ vx*wx4 
  wy4 ~~ vy*wy4 
  wx5 ~~ vx*wx5
  wy5 ~~ vy*wy5
  
  # Constrain grand means over time
  x1 + x2 + x3 + x4 + x5 ~ mx*1
  y1 + y2 + y3 + y4 + y5 ~ my*1
'
RICLPM5.fit <- lavaan(RICLPM5, 
  data = dat, 
  missing = 'ML', 
  meanstructure = T, 
  int.ov.free = T
) 
summary(RICLPM5.fit, standardized = T)
```


#3 Constrain standardized lagged effects

Rather than imposing constraints on the unstandardized auto-regressive and cross-lagged relations over time, we can impose constraints on the standardized lagged effects, as explained in the FAQ. Compared to the previous model (with constraints over time on the unstandardized lagged effects), there are multiple changes to the syntax. First, we freely estimate the factor loadings that link the observed variables to the within-components (by premultiplying with NA), rather than fixing them to 1:
```{r}
# Create within-components with freely estimated factor loadings
  wx1 =~ NA*x1
  wx2 =~ NA*x2
  wx3 =~ NA*x3 
  wx4 =~ NA*x4
  wx5 =~ NA*x5
  wy1 =~ NA*y1
  wy2 =~ NA*y2
  wy3 =~ NA*y3
  wy4 =~ NA*y4
  wy5 =~ NA*y5
```

Second, we set the variances of within-components at first wave to 1, and label the covariance (now also the correlation) between them:
```{r}
# Set variances of within-components at first wave to 1
  wx1 ~~ 1*wx1
  wy1 ~~ 1*wy1
  
  # Estimate correlation between within-components at first wave
  wx1 ~~ cor1*wy1 
```

Third, we give the residual variance and covariances between the within-component each a unique label:
```{r}
 # Compute correlations of within-components at each wave
  cor2 := a*c + b*d + a*d*cor1 + b*c*cor1 + rcov2
  cor3 := a*c + b*d + a*d*cor2 + b*c*cor2 + rcov3
  cor4 := a*c + b*d + a*d*cor3 + b*c*cor3 + rcov4
 
  # Contrain residual variances of within-components such that variance of each 
  # within-component equals 1
  rvx2 == 1 - (a*a + b*b + 2*a*b*cor1)
  rvy2 == 1 - (c*c + d*d + 2*c*d*cor1)
  rvx3 == 1 - (a*a + b*b + 2*a*b*cor2)
  rvy3 == 1 - (c*c + d*d + 2*c*d*cor2)
  rvx4 == 1 - (a*a + b*b + 2*a*b*cor3)
  rvy4 == 1 - (c*c + d*d + 2*c*d*cor3)
  rvx5 == 1 - (a*a + b*b + 2*a*b*cor4)
  rvy5 == 1 - (c*c + d*d + 2*c*d*cor4)
```

These steps ultimately result in the below syntax. Comparing the constrainted unstandardized and constrained standardized results shows that these are now equivalent.

```{r}
RICLPM6 <- '
  # Create between components (random intercepts)
  RIx =~ 1*x1 + 1*x2 + 1*x3 + 1*x4 + 1*x5
  RIy =~ 1*y1 + 1*y2 + 1*y3 + 1*y4 + 1*y5
  
  # Create within-components with freely estimated factor loadings
  wx1 =~ NA*x1
  wx2 =~ NA*x2
  wx3 =~ NA*x3 
  wx4 =~ NA*x4
  wx5 =~ NA*x5
  wy1 =~ NA*y1
  wy2 =~ NA*y2
  wy3 =~ NA*y3
  wy4 =~ NA*y4
  wy5 =~ NA*y5
  
  # Estimate lagged effects between within-person centered variables 
  # (constrained)
  wx2 ~ a*wx1 + b*wy1 
  wy2 ~ c*wx1 + d*wy1
  wx3 ~ a*wx2 + b*wy2
  wy3 ~ c*wx2 + d*wy2
  wx4 ~ a*wx3 + b*wy3
  wy4 ~ c*wx3 + d*wy3
  wx5 ~ a*wx4 + b*wy4
  wy5 ~ c*wx4 + d*wy4
  
  # Label residual covariances
  wx2 ~~ rcov2*wy2
  wx3 ~~ rcov3*wy3
  wx4 ~~ rcov4*wy4 
  wx5 ~~ rcov5*wy5
  
  # Estimate correlation between within-components at first wave
  wx1 ~~ cor1*wy1 
  
  # Estimate variance and covariance of random intercepts
  RIx ~~ RIx
  RIy ~~ RIy
  RIx ~~ RIy
  
  # Set variances of within-components at first wave to 1
  wx1 ~~ 1*wx1
  wy1 ~~ 1*wy1
  
  # Label residual variances
  wx2 ~~ rvx2*wx2 
  wy2 ~~ rvy2*wy2 
  wx3 ~~ rvx3*wx3 
  wy3 ~~ rvy3*wy3 
  wx4 ~~ rvx4*wx4 
  wy4 ~~ rvy4*wy4 
  wx5 ~~ rvx5*wx5
  wy5 ~~ rvy5*wy5
  
  # Constrain grand means over time
  x1 + x2 + x3 + x4 + x5 ~ mx*1
  y1 + y2 + y3 + y4 + y5 ~ my*1
  
  # Compute correlations of within-components at each wave
  cor2 := a*c + b*d + a*d*cor1 + b*c*cor1 + rcov2
  cor3 := a*c + b*d + a*d*cor2 + b*c*cor2 + rcov3
  cor4 := a*c + b*d + a*d*cor3 + b*c*cor3 + rcov4
 
  # Contrain residual variances of within-components such that variance of each 
  # within-component equals 1
  rvx2 == 1 - (a*a + b*b + 2*a*b*cor1)
  rvy2 == 1 - (c*c + d*d + 2*c*d*cor1)
  rvx3 == 1 - (a*a + b*b + 2*a*b*cor2)
  rvy3 == 1 - (c*c + d*d + 2*c*d*cor2)
  rvx4 == 1 - (a*a + b*b + 2*a*b*cor3)
  rvy4 == 1 - (c*c + d*d + 2*c*d*cor3)
  rvx5 == 1 - (a*a + b*b + 2*a*b*cor4)
  rvy5 == 1 - (c*c + d*d + 2*c*d*cor4)
'
RICLPM6.fit <- lavaan(RICLPM6, 
  data = dat, 
  missing = 'ML', 
  meanstructure = T, 
  int.ov.free = T
) 
summary(RICLPM6.fit, standardized = F)
summary(RICLPM6.fit, standardized = T)
```

