---
title: "Thesis Data Cleaning"
author: "Yeon Suh"
date: "`r Sys.Date()`"
output: html_document
---


```{r read LB data}
LB22 <- read_sas("H22LB_R.sas7bdat")
LB20 <- read_sas("H20LB_R.sas7bdat")
LB18 <- read_sas("H18LB_R.sas7bdat")

fn <- "H16LB_R.da" #call da dataset
sas.input <- "H16LB_R.sas" #call sas dataset
LB16 <- read.SAScii( fn , sas.input ) #read both

fn <- "H14LB_R.da" #call da dataset
sas.input <- "H14LB_R.sas" #call sas dataset
LB14 <- read.SAScii( fn , sas.input ) #read both

fn <- "H12LB_R.da" #call da dataset
sas.input <- "H12LB_R.sas" #call sas dataset
LB12 <- read.SAScii( fn , sas.input ) #read both
```

```{r}
lbeftf1 <- LB22 %>%
  full_join(LB18, by = c("HHID", "PN")) %>%
  full_join(LB14, by = c("HHID", "PN"))

lbeftf2 <- LB20 %>%
  full_join(LB16, by = c("HHID", "PN")) %>%
  full_join(LB12, by = c("HHID","PN"))

saveRDS(lbeftf1, "lbeftf1.rds")
saveRDS(lbeftf2, "lbeftf2.rds")
```

```{r}
library(psych)
library(dplyr)
library(lavaan)
library(SAScii) 
library(haven)
library(skimr)
```

```{r read TRACKER}
#tracker = read_sas("TRK2022TR_R.sas7bdat")
#saveRDS(tracker, "tracker.RDS")
tracker = readRDS("tracker.RDS")
```

EFTFASSIGN               RESPONDENTS ASSIGNMENT FOR EFTF INTERVIEWS
         Section: TR    Level: Respondent      Type: Numeric    Width: 1   Decimals: 0

         .................................................................................
         19782           1.  To receive enhanced face to face interview in 2006, 2010,
                             2014, 2018, and 2022
         18523           2.  To receive enhanced face to face interview in 2008, 2012,
                             2016, 2020, and 2024
          8461           9.  Removed from sample prior to the 2006 data collection

```{r}
eftf1 = tracker[tracker$EFTFASSIGN==1,]
eftf2 = tracker[tracker$EFTFASSIGN==2,]
```

```{r}
eftf1 = eftf1[c("HHID","PN",
                #HOUSEHOLD NUMBER, PERSON NUMBER
                "SEX","RACE","SCHLYRS","OAGE",
                #GENDER, RACE, SCHOLL YEARS, BASELINE AGE
                "OALIVE","QALIVE","SALIVE",
                #VITAL STATUS
                "OINSAMP","QINSAMP","SINSAMP",
                #IN SAMPLE
                "OCOUPLE","QCOUPLE","SCOUPLE",
                #COUPLE STATUS:married or living with a partner as if married
                #1.YES 5.NO 
                "OLBCOMP","QLBCOMP","SLBCOMP",
                #DID COMPLETE LEAVE BEHIND QUESTIONNAIRE?
                "OLIVARR","QLIVARR","SLIVARR",
                #LIVING ARRANGEMENT
                "OMARST","QMARST","SMARST",
                #MARITAL STATUS
                # 1.MARRIED 2.SEPARATED/DIVORCED 
                # 3.WIDOWED 4.NEVER MARRIED 5.UNKNOWN
                "ONURSHM","QNURSHM","SNURSHM"
                #NURSINGHOME 
                )]

eftf2 = eftf2[c("HHID","PN",
                "SEX","RACE","SCHLYRS","NAGE",
                "NALIVE","PALIVE","RALIVE",
                "NINSAMP","PINSAMP","RINSAMP",
                "NCOUPLE","PCOUPLE","RCOUPLE",
                "NLBCOMP","PLBCOMP","RLBCOMP",
                "NLIVARR","PLIVARR","RLIVARR",
                "NMARST","PMARST","RMARST",
                "NNURSHM","PNURSHM","RNURSHM"
                )]
```


```{r}
#PARTICIPANT IN SAMPLE AT BASELINE
eftf1 = subset(eftf1, OINSAMP == 1) #N = 11,737
eftf2 = subset(eftf2, NINSAMP == 1)#N = 12,190

#PARTICIPANT ALIVE AT BASELINE
eftf1 = subset(eftf1, OALIVE == 1 | OALIVE == 2)
#PARTICIPANT MUST BE 50 YEARS OR OLDER
eftf1 = subset(eftf1, OAGE >= 50 & OAGE != 999)

eftf2 = subset(eftf2, NALIVE == 1 | NALIVE == 2)
eftf2 = subset(eftf2, NAGE >= 50 & NAGE !=999)

eftf1 #9,325
eftf2 #9,727
```

```{r}
randeftf1 = rand[c("HHID","PN",
                  #PROXY 0.NO, 1.YES
              "R12PROXY","R14PROXY","R16PROXY",
              #MOBILITY: 0-5 SUM OF DIFFICULTY IN WALKING/CLIMBING STAIRS
              "R12MOBILA","R14MOBILA","R16MOBILA",
              #TOTAL HOUSEHOLD WEALTH
              "H12ATOTB","H14ATOTB","H16ATOTB",
              #NUMBER OF PEOPLE LIVING IN THE HOUSEHOLD
              "H12HHRES","H14HHRES","H16HHRES",
              #NUMBER OF LIVING CHILDREN
              "H12CHILD","H14CHILD","H16CHILD")]

randeftf2 = rand[c("HHID","PN",
                  #PROXY 0.NO, 1.YES
              "R11PROXY","R13PROXY","R15PROXY",
              #MOBILITY: 0-5 SUM OF DIFFICULTY IN WALKING/CLIMBING STAIRS
              "R11MOBILA","R13MOBILA","R15MOBILA",
              #TOTAL HOUSEHOLD WEALTH
              "H11ATOTB","H13ATOTB","H15ATOTB",
              #NUMBER OF PEOPLE LIVING IN THE HOUSEHOLD
              "H11HHRES","H13HHRES","H15HHRES",
              #NUMBER OF LIVING CHILDREN
              "H11CHILD","H13CHILD","H15CHILD")]
```

```{r}
eftf1 <- eftf1 %>%
  left_join(randeftf1, by = c("HHID", "PN"))
eftf1

eftf2<- eftf2 %>%
  left_join(randeftf2, by = c("HHID", "PN"))
eftf2
```

#LBQ Variables

```{r LB data}
lbeftf1 = readRDS("lbeftf1.rds")
lbeftf2 = readRDS("lbeftf2.rds")
```

```{r LB Loneliness 11scale}
reverse_lonely <- function(x) 4 - x

make_loneliness <- function(data, prefix, wave, item_num = "019") {
  items <- paste0(prefix, "LB", item_num, LETTERS[1:11])

  rev_items <- paste0(prefix, "LB", item_num, c("A", "B", "C", "E"))

  data %>%
    mutate(
      across(all_of(rev_items), reverse_lonely, .names = "rev_{.col}"),

      "loneliness11.{wave}" := ifelse(
        #IF MISSING MORE THAN 5 ITEMS => NA
        rowSums(is.na(across(all_of(items)))) > 5,
        NA_real_,
        rowMeans(
          cbind(
            across(all_of(paste0("rev_", rev_items))),
            across(all_of(setdiff(items, rev_items)))
          ),
          na.rm = TRUE
        )
      )
    )
}

lbeftf1 <- lbeftf1 %>%
  make_loneliness("O", 1, item_num = "019") %>%
  make_loneliness("Q", 2, item_num = "019") %>%
  make_loneliness("S", 3, item_num = "019")

lbeftf2 <- lbeftf2 %>%
  make_loneliness("N", 1, item_num = "020") %>%
  make_loneliness("P", 2, item_num = "019") %>%
  make_loneliness("R", 3, item_num = "019")

lbeftf1 %>%
  select(starts_with("loneliness")) %>%
  summary()

lbeftf2 %>%
  select(starts_with("loneliness")) %>%
  summary()

lbeftf1[c("loneliness11.1","OLB019A","OLB019B","OLB019C",
          "OLB019D","OLB019E","OLB019F","OLB019G","OLB019H","OLB019I",
          "OLB019J","OLB019K")]

summary(lbeftf1$loneliness11.1)
summary(lbeftf2$loneliness11.1)

```


```{r}
lbeftf1 <- lbeftf1 %>%
  mutate(
    haschild1 = ifelse(OLB006 == 1, 1, ifelse(OLB006 == 5, 0, NA)),
    haschild2 = ifelse(QLB006 == 1, 1, ifelse(QLB006 == 5, 0, NA)),
    haschild3 = ifelse(SLB006 == 1, 1, ifelse(SLB006 == 5, 0, NA)),
    hasfriend1 = ifelse(OLB014 == 1, 1, ifelse(OLB014 == 5, 0, NA)),
    hasfriend2 = ifelse(QLB014 == 1, 1, ifelse(QLB014 == 5, 0, NA)),
    hasfriend3 = ifelse(SLB014 == 1, 1, ifelse(SLB014 == 5, 0, NA)),
    livewithsp1 = ifelse(OLB003 == 1, 1, ifelse(OLB003 == 5, 0, NA)),
    livewithsp2 = ifelse(QLB003 == 1, 1, ifelse(QLB003 == 5, 0, NA)),
    livewithsp3 = ifelse(SLB003 == 1, 1, ifelse(SLB003 == 5, 0, NA))
  )

lbeftf2 <- lbeftf2 %>%
  mutate(
    haschild1 = ifelse(NLB007 == 1, 1, ifelse(NLB007 == 5, 0, NA)),
    haschild2 = ifelse(PLB006 == 1, 1, ifelse(PLB006 == 5, 0, NA)),
    haschild3 = ifelse(RLB006 == 1, 1, ifelse(RLB006 == 5, 0, NA)),
    hasfriend1 = ifelse(NLB015 == 1, 1, ifelse(NLB015 == 5, 0, NA)),
    hasfriend2 = ifelse(PLB014 == 1, 1, ifelse(PLB014 == 5, 0, NA)),
    hasfriend3 = ifelse(RLB014 == 1, 1, ifelse(RLB014 == 5, 0, NA)),
    livewithsp1 = ifelse(NLB004 == 1, 1, ifelse(NLB004 == 5, 0, NA)),
    livewithsp2 = ifelse(PLB003 == 1, 1, ifelse(PLB003 == 5, 0, NA)),
    livewithsp3 = ifelse(RLB003 == 1, 1, ifelse(RLB003 == 5, 0, NA))
  )

#lbeftf1[c("haschild1", "OLB006")]
#lbeftf2[c("haschild1", "NLB007")]
```

```{r support & strain}
make_social_support <- function(data, prefix, wave,
                                spouse_num = "004",
                                child_num  = "007",
                                friend_num = "015") {

  sp_pos <- paste0(prefix, "LB", spouse_num, c("A", "B", "C"))
  sp_neg <- paste0(prefix, "LB", spouse_num, c("D", "E", "F", "G"))

  ch_pos <- paste0(prefix, "LB", child_num, c("A", "B", "C"))
  ch_neg <- paste0(prefix, "LB", child_num, c("D", "E", "F", "G"))

  fr_pos <- paste0(prefix, "LB", friend_num, c("A", "B", "C"))
  fr_neg <- paste0(prefix, "LB", friend_num, c("D", "E", "F", "G"))

  data %>%
    mutate(across(all_of(c(sp_pos, sp_neg, ch_pos, ch_neg, fr_pos, fr_neg)),
                  ~ 5 - ., .names = "rev_{.col}")) %>%
    mutate(
      "spsupport{wave}" := ifelse(rowSums(is.na(across(all_of(sp_pos)))) > 1, NA_real_,
                                  rowMeans(across(all_of(paste0("rev_", sp_pos))), na.rm = TRUE)),
      "spstrain{wave}"  := ifelse(rowSums(is.na(across(all_of(sp_neg)))) > 2, NA_real_,
                                  rowMeans(across(all_of(paste0("rev_", sp_neg))), na.rm = TRUE)),

      "chsupport{wave}" := ifelse(rowSums(is.na(across(all_of(ch_pos)))) > 1, NA_real_,
                                  rowMeans(across(all_of(paste0("rev_", ch_pos))), na.rm = TRUE)),
      "chstrain{wave}"  := ifelse(rowSums(is.na(across(all_of(ch_neg)))) > 2, NA_real_,
                                  rowMeans(across(all_of(paste0("rev_", ch_neg))), na.rm = TRUE)),

      "frsupport{wave}" := ifelse(rowSums(is.na(across(all_of(fr_pos)))) > 1, NA_real_,
                                  rowMeans(across(all_of(paste0("rev_", fr_pos))), na.rm = TRUE)),
      "frstrain{wave}"  := ifelse(rowSums(is.na(across(all_of(fr_neg)))) > 2, NA_real_,
                                  rowMeans(across(all_of(paste0("rev_", fr_neg))), na.rm = TRUE))
    )
}

lbeftf1 <- lbeftf1 %>%
  make_social_support("O", 1) %>%
  make_social_support("Q", 2) %>%
  make_social_support("S", 3)

lbeftf2 <- lbeftf2 %>%
  make_social_support("N", 1, spouse_num = "005", child_num = "008", friend_num = "016") %>%
  make_social_support("P", 2) %>%
  make_social_support("R", 3)


#SANITY CHECK
lbeftf1[c("frsupport1","frsupport2","frsupport3")]
lbeftf2[c("frsupport1","frsupport2","frsupport3")]

lbeftf1[c("frsupport1","OLB015A","OLB015B","OLB015C",
          "frstrain1","OLB015D","OLB015E","OLB015F","OLB015G")]

```

                      Q09. # CHILDREN CLOSE RELATIONSHIP
         Section: LB    Level: Respondent      Type: Numeric    Width: 2   Decimals: 0

         How many of your children would you say you have a close relationship with?
         (Please write a number in the box.)


```{r close chilren}
lbeftf2$numclosechild1 = lbeftf2$NLB010
lbeftf1$numclosechild1 = lbeftf1$OLB009
lbeftf2$numclosechild2 = lbeftf2$PLB009
lbeftf1$numclosechild2 = lbeftf1$QLB009
lbeftf2$numclosechild3 = lbeftf2$RLB009
lbeftf1$numclosechild3 = lbeftf1$SLB009

#Cap at 3
lbeftf1 <- lbeftf1 %>%
  mutate(
    closechild1 = pmin(numclosechild1, 3),
    closechild2 = pmin(numclosechild2, 3),
    closechild3 = pmin(numclosechild3, 3)
  )

lbeftf2 <- lbeftf2 %>%
  mutate(
    closechild1 = pmin(numclosechild1, 3),
    closechild2 = pmin(numclosechild2, 3),
    closechild3 = pmin(numclosechild3, 3)
  )
```


```{r lbeftf needed variables only}
cleanlbeftf1 = lbeftf1[c("HHID","PN",
                         "haschild1","haschild2","haschild3",
                         "hasfriend1","hasfriend2","hasfriend3",
                         "livewithsp1","livewithsp2","livewithsp3",
                         "spsupport1","spsupport2","spsupport3",
                         "chsupport1","chsupport2","chsupport3",
                         "frsupport1","frsupport2","frsupport3",
                         "numclosechild1","numclosechild2","numclosechild3",
                         "closechild1","closechild2","closechild3",
                         "loneliness11.1","loneliness11.2","loneliness11.3"
                         )]
cleanlbeftf2 = lbeftf2[c("HHID","PN",
                         "haschild1","haschild2","haschild3",
                         "hasfriend1","hasfriend2","hasfriend3",
                         "livewithsp1","livewithsp2","livewithsp3",
                         "spsupport1","spsupport2","spsupport3",
                         "chsupport1","chsupport2","chsupport3",
                         "frsupport1","frsupport2","frsupport3",
                         "numclosechild1","numclosechild2","numclosechild3",
                         "closechild1","closechild2","closechild3",
                         "loneliness11.1","loneliness11.2","loneliness11.3"
                         )]
```



```{r}
eftf1 <- eftf1 %>%
  left_join(cleanlbeftf1, by = c("HHID", "PN"))
eftf1

eftf2<- eftf2 %>%
  left_join(cleanlbeftf2, by = c("HHID", "PN"))
eftf2
```

```{r}
names(eftf1)
names(eftf2)
```

```{r}
eftf1cleaned <- eftf1 %>%
  rename(
    gender = SEX, race = RACE, edu = SCHLYRS, baselineage = OAGE,
    alive1 = OALIVE, alive2 = QALIVE, alive3 = SALIVE,
    lbcomp1 = OLBCOMP, lbcomp2 = QLBCOMP, lbcomp3 = SLBCOMP,
    couple1 = OCOUPLE, couple2 = QCOUPLE, couple3 = SCOUPLE,
    livarr1 = OLIVARR, livarr2 = QLIVARR, livarr3 = SLIVARR,
    mstat1 = OMARST, mstat2 = QMARST, mstat3 = SMARST,
    numchild1 = H12CHILD, numchild2 = H14CHILD, numchild3 = H16CHILD,
    proxy1 = R12PROXY, proxy2 = R14PROXY, proxy3 = R16PROXY,
    mobility1 = R12MOBILA, mobility2 = R14MOBILA, mobility3 = R16MOBILA,
    wealth1 = H12ATOTB, wealth2 = H14ATOTB, wealth3 = H16ATOTB,
    hhsize1 = H12HHRES, hhsize2 = H14HHRES, hhsize3 = H16HHRES,
    nursing1 = ONURSHM, nursing2 = QNURSHM, nursing3=SNURSHM
  )

eftf2cleaned <- eftf2 %>%
  rename(
    gender = SEX, race = RACE, edu = SCHLYRS, baselineage = NAGE,
    alive1 = NALIVE, alive2 = PALIVE, alive3 = RALIVE,
    couple1 = NCOUPLE, couple2 = PCOUPLE, couple3 = RCOUPLE,
    lbcomp1 = NLBCOMP, lbcomp2 = PLBCOMP, lbcomp3 = RLBCOMP, 
    livarr1 = NLIVARR, livarr2 = PLIVARR, livarr3 = RLIVARR,
    mstat1 = NMARST, mstat2 = PMARST, mstat3 = RMARST,
    numchild1 = H11CHILD, numchild2 = H13CHILD, numchild3 = H15CHILD,
    proxy1 = R11PROXY, proxy2 = R13PROXY, proxy3 = R15PROXY,
    mobility1 = R11MOBILA, mobility2 = R13MOBILA, mobility3 = R15MOBILA,
    wealth1 = H11ATOTB, wealth2 = H13ATOTB, wealth3 = H15ATOTB,
    hhsize1 = H11HHRES, hhsize2 = H13HHRES, hhsize3 = H15HHRES,
    nursing1 = NNURSHM, nursing2 = PNURSHM, nursing3=RNURSHM
  )
```

```{r}
setdiff(names(eftf1cleaned), names(eftf2cleaned))
setdiff(names(eftf2cleaned), names(eftf1cleaned))
```

```{r}
#STACK EFTF1&2
stacked <- bind_rows(eftf1cleaned, eftf2cleaned)

#ERASE NOT NEEDED COL
stacked$OINSAMP = NULL
stacked$QINSAMP = NULL
stacked$SINSAMP = NULL
stacked$NINSAMP = NULL
stacked$PINSAMP = NULL
stacked$RINSAMP = NULL
stacked
```

```{r CLEANING}
#CLOSE CHILD RATIO
stacked$ccratio1 = (stacked$numclosechild1/stacked$numchild1)
stacked$ccratio2 = (stacked$numclosechild2/stacked$numchild2)
stacked$ccratio3 = (stacked$numclosechild3/stacked$numchild3)

#CAP NUMCHILD AT 5
stacked <- stacked %>%
  mutate(
    numchild1 = pmin(numchild1, 5),
    numchild2 = pmin(numchild2, 5),
    numchild3 = pmin(numchild3, 5)
  )

#HOUSEHOLD SIZE
#Cap at 3
stacked <- stacked %>%
  mutate(
    hhsize1 = pmin(hhsize1, 3),
    hhsize2 = pmin(hhsize2, 3),
    hhsize3 = pmin(hhsize3, 3)
  )

#GENDER
stacked <- stacked %>%
  mutate(
    gender = case_when(
      gender == 1 ~ 0, # 0.MALE 
      gender == 2 ~ 1, # 1.FEMALE
      TRUE ~ NA_real_
    )
  )

#WHITE/NON-WHITE
stacked$white <- ifelse(
  stacked$race == 1, 1, # 0.NON-WHITE 1.WHITE
  ifelse(stacked$race %in% c(2, 7), 0, NA)
)

#QUINTILE WEALTH
stacked <- stacked %>%
  mutate(
    wealth1 = ntile(wealth1, 5),
    wealth2 = ntile(wealth2, 5),
    wealth3 = ntile(wealth3, 5)
  )

#SIMPLIFIY MARITAL STATUS INTO 0.NOT MARRIED 1.MARRIED
#NOT MARRIED INCLUDING NEVER MARRIED, WIDOWED, DIVORCED, ETC.
stacked <- stacked %>%
  mutate(
    married1 = ifelse(is.na(mstat1), NA,
                      ifelse(mstat1 == 1, 1, 0)),

    married2 = ifelse(is.na(mstat2), NA,
                      ifelse(mstat2 == 1, 1, 0)),

    married3 = ifelse(is.na(mstat3), NA,
                      ifelse(mstat3 == 1, 1, 0))
  )

#VITAL STATUS
stacked <- stacked %>%
  mutate(
    deceased = ifelse(
      (!is.na(alive1) & !(alive1 %in% c(1, 2))) |
      (!is.na(alive2) & !(alive2 %in% c(1, 2))) |
      (!is.na(alive3) & !(alive3 %in% c(1, 2))), 1, 0
    )
  )

#COUPLE STATUS
stacked <- stacked %>%
  mutate(
    couple1 = ifelse(couple1 == 1, 1,
                     ifelse(couple1 == 5, 0, NA)),

    couple2 = ifelse(couple2 == 1, 1,
                     ifelse(couple2 == 5, 0, NA)),

    couple3 = ifelse(couple3 == 1, 1,
                     ifelse(couple3 == 5, 0, NA))
  )

#SCHOOL YEARS
stacked$edu[stacked$edu == 99] <- NA

table(stacked$alive1, useNA = "ifany")
table(stacked$alive2, useNA = "ifany")
table(stacked$alive3, useNA = "ifany")

stacked


#19,052
saveRDS(stacked, "thesisdata_stacked.rds")
```


```{r}
alive <- stacked %>%
  filter(
    alive1 %in% c(1, 2) & alive2 %in% c(1, 2) & alive3 %in% c(1, 2) &
    !(alive1 == 2 & alive2 == 2 & alive3 == 2)
  )

alive <- alive %>%
  filter(
    (proxy1 == 0 | is.na(proxy1)) &
    (proxy2 == 0 | is.na(proxy2)) &
    (proxy3 == 0 | is.na(proxy3))
  )

alive = alive %>%
  filter(
    (nursing1 == 5 | is.na(nursing1)) &
      (nursing2 == 5 | is.na(nursing2)) &
      (nursing3 == 5 | is.na(nursing3))
  )

alive <- alive %>%
  filter(
    !(lbcomp1 == 5 &
      lbcomp2 == 5 &
      lbcomp3 == 5)
  )

alive$proxy1 = NULL
alive$proxy2 = NULL
alive$proxy3 = NULL
alive$alive1 = NULL
alive$alive2 = NULL
alive$alive3 = NULL
alive$nursing1 = NULL
alive$nursing2 = NULL
alive$nursing3 = NULL
alive
#8,488
```

```{r}
saveRDS(alive, "thesisdata_alive.rds")
```


```{r}
deceased <- stacked %>%
  filter(
    (!is.na(alive1) & !(alive1 %in% c(1, 2))) |
    (!is.na(alive2) & !(alive2 %in% c(1, 2))) |
    (!is.na(alive3) & !(alive3 %in% c(1, 2)))
  )
deceased
#5,019
saveRDS(deceased, "thesisdata_deceased.rds")
```























